---
title: "Biomedical signal processing and visualisation using R"
subtitle: "VII Encontro Nacional de Engenharia Biomecânica"
author: "Prof. Adriano de Oliveira Andrade, PhD - adriano@ufu.br"
date: "01 June, 2022"
output: 
# https://bookdown.org/yihui/rmarkdown-cookbook/cross-ref.html
       bookdown::html_document2:
         toc: true 
         #toc_float:
         # toc_collapsed: true
         toc_depth: 6
         theme: readable
         #https://bookdown.org/yihui/rmarkdown/html-document.html#custom-css
         highlight: haddock 
         fig_caption: yes
         number_sections: true
         css: style.css


---

<head>
    <link rel="stylesheet" href="https://cdnjs.cloudflare.com/ajax/libs/font-awesome/4.7.0/css/font-awesome.min.css"> 
</head>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Biomedical signal visualisation

## Static visualising with R

The first thing to do is open the data file. The data set in this example is derived from inertial data, such as accelerometer, magnetometer, and gyroscope. The information was saved in an ``.xlsx`` (Excel) file.

```{r warning=FALSE}
library(openxlsx)

df <- openxlsx::read.xlsx("V16C1RCC92.xlsx") # load inertial data from Excel file

names(df) # display the variable names

```

The names of the variables can be improved to make graphic generation easier.


```{r warning=FALSE}
library(stringr)

aa <- names(df) # assigning the variable names to a variable

# identifying and replacing patterns in the names of the variables
aa <- str_replace(string = aa, pattern = c("\\["), replacement = "") 
aa <- str_replace(string = aa, pattern = "\\]", replacement = "") 
aa <- str_replace(string = aa, pattern = "\\.", replacement = "")

names(df) <- aa # assigning the altered names to the variables

names(df)

```

As the data is now available in the global environment it is possible to visualize it by using static and interactive graphics. Let's start looking on how to plot the signal in a static way, which is very suitable for publications in journals, theses and other static materials.

```{r warning=FALSE, fig.height=1.5, fig.width=10}
library(ggplot2)

gg <- ggplot() # create the graphic object
gg <- gg + ggplot2::geom_line(data = df, 
                              mapping = aes(x = Time, y = G1X)) # adding layers to the graphic object
gg # display the graphic on the screen

gg <- gg + theme_minimal() #changing the appearance of the theme 
gg # display the graphic on the screen

gg <- gg + ggplot2::ggtitle(label = "gyroscope data - sensor 1")

gg # display the graphic on the screen

```

If you wish to plot several biomedical time series in the same graphic or in distinct subpanels the data should be first reorganized to easy plotting the data.

```{r warning=FALSE}

library(tidyr)

df.long <- pivot_longer(data = df, cols = (2:length(df)), 
                        names_to = "variable",
                        values_to = "value")
head(df.long, n = 10L)

# converting the variable names from string to factors
# to sort out the variables according to their levels
df.long$variable <- factor(df.long$variable, 
                           levels = df.long$variable %>% unique()) 


```

Once the data is organized it can be plotted by using the ``ggplot2`` package.


```{r warning=FALSE, fig.width=10, fig.height=15}
g <- ggplot(data = df.long,
            mapping = aes(x = Time, y = value)) +
  geom_line() + 
  facet_wrap( ~ variable, ncol = 3, scales = "free_y") + 
  theme_light() + labs(y = "", x = "time (s)")

g

```

If you want to plot several waveforms on the same graph, you can choose the signals you want and plot them as shown below.

```{r fig.height=2, fig.width=10, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)

df.filtered <- df.long %>% filter(variable == "G1X" |
                                    variable == "G1Y" |
                                    variable == "G1Z")

g <- ggplot(data = df.filtered,
            mapping = aes(x = Time,
                          y = value,
                          color = interaction(variable))) +
  geom_line() + theme_minimal() + labs(color = "signal",
                                       x = "time (s)",
                                       y = "angular velocity (º/s)") 
g


```

Several waveforms can be grouped and plotted in panels based on specific criteria.

```{r fig.height=3, fig.width=10, message=FALSE, warning=FALSE}
df.filtered <- df.long %>% filter(variable == "G1X" |
                                    variable == "G1Y" |
                                    variable == "G1Z" |
                                    variable == "A1X" |
                                    variable == "A1Y" |
                                    variable == "A1Z" |
                                    variable == "M1X" |
                                    variable == "M1Y" |
                                    variable == "M1Z")

df.filtered$sensor <- str_sub(df.filtered$variable, 1, 1)

g <- ggplot(data = df.filtered,
            mapping = aes(x = Time,
                          y = value,
                          color = interaction(variable))) +
  geom_line() +
  facet_wrap( ~ sensor, ncol = 3, scales = "free_y",
              strip.position = "left", 
              labeller = as_labeller(c(A = "linear acceleration (g)", 
                                       G = "angular velocity (º/s)",
                                       M = "magnetic field (Gauss)"))) +
  
  theme_minimal() + labs(color = "signal",
                         x = "time (s)",
                         y = "") +
  theme(strip.background = element_blank(),
        strip.placement = "outside")
g
```

Static figures can be saved in vectorial and several other formats.

```{r}

ggsave(filename = "figure1.pdf", plot = g, width = 160,
       height = 75, units = "mm")

ggsave(filename = "figure1.png", plot = g, width = 160,
       height = 75, units = "mm")

ggsave(filename = "figure1.tiff", plot = g, width = 160,
       height = 75, units = "mm")

```

**Now, it is time for action!**

> Generate a static graph with 'ggplot2', containing a panel for the X axis of each sensor type (i.e., accelerometer, gyroscope, and magnetometer). In each panel, two time-series should be plotted for each IMU (i.e. IMU1 and IMU2), and the time-series of IMU1 should be plotted in black while those for IMU2 in red. Don't forget to include the axes' names and units of measurement!

## Interactive visualizaton with R

When inspecting and processing biomedical signals, interactive data visualization becomes very important and can be fundamental for understanding information, navigating the signal, paying attention to details and identifying trends, discontinuities, or outliers, because this type of data visualization allows us to zoom in on periods of interest, highlight points, and mark events that help in data analysis. With the ``dygraphs`` package in R, we bring our graphics to life. It's simple to create an dynamic graphic with this package and the code below does just that!

```{r message=FALSE, warning=FALSE, error=FALSE, fig.width=9, fig.height=1.5}

library(dygraphs)

df.filtered <- df %>% 
  dplyr::select(Time, G1Y) # selecting the Time and G1Y columns

dygraph(df.filtered) # generating the dynamic graph

```

Although we can zoom in on signal segments (by clicking and dragging) and re-view the entire signal with a double click, we still can't navigate along the signal. When we are interested in navigating along the signal or still paying attention to time periods, we can use the ``dyRangeSelector()`` function. 

```{r message=FALSE, warning=FALSE, error=FALSE, fig.width=9, fig.height=1.5}

dygraph(df.filtered) %>% # generating the dynamic graph
  dyRangeSelector(dateWindow = c(10,17)) # adding a range selector (rectangular window of 7 s)

```

In addition, the ``dyEvent()`` function is used to specify the times when events occurred. Event marking is particularly interesting when we believe that the event has probably influenced a noticeable change in the time series and should be observed. In other hand, the ``dyShading()`` function allows us to highlight a period or several periods in the time series data. Additionally, you can specify the color of the series lines and other graph styling by using the ``dyOptions()`` function.

```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center',fig.width=9, fig.height=3.5}

df.filtered <- df %>% 
  dplyr::select(Time, G1X, G1Y, G1Z) # selecting the Time and gyroscope signals of IMU1

dygraph(df.filtered, xlab = "time (s)", ylab = "angular velocity (º/s)") %>% # generating the dynamic graph
  dyRangeSelector(dateWindow = c(15,20)) %>% # adding a range selector
  dyOptions(colors = c("#ae0001","#d3a625","#2a623d")) %>% # adding color palette
  dyShading(from = "17.02", to = "17.34", color = "#cee6fd") %>% # adding shadow
  dyShading(from = "19.18", to = "19.54", color = "#fff5da") %>% # adding shadow
  dyEvent(c("17.02", "17.34"), c("start", "end"), labelLoc = "top") # adding event

```

# Pre-processing biomedical signals

## Signal windowing

Signal windowing is the process of segmenting a signal into sections in order to facilitate analysis and improve the spectral properties of the sampled signal. In general, different types of windows can be used for windowing (i.e., rectangular window, Hanning window, Hamming window, and Kaiser-Bessel window, among other types of window).

However, defining a window size given in units of time or number of samples is a significant challenge in practice. We can't consistently segment the signal if we choose a size that is too small or too large. Thus, knowledge of the investigated phenomenon, sampling frequency, and signal duration are required to define the size of the window.

Furthermore, signal windowing can be done with or without overlap. When there is overlap, a window intersects with the adjacent window at some time point. On the other hand, when there is no overlap, the start time of a window does not overlap with the time of the adjacent window. When we want to investigate signal behavior over time using features, we usually use unoverlaped rectangular windows. This type of window is straightforward and easy to implement in R. 

We will window the EMG signal of channel 2 using non-overlapping rectangular windows of 500 ms. Thus, knowing the sampling frequency of the signal and its duration, we can define the start and end indexes of each window.

For the sampling frequency ($f_s$), it is enough to consider the inverse of temporal resolution ($\Delta t$), that is,

$$f_s = \frac{1}{\Delta t}$$

the inverse of the time difference between two consecutive samples.

```{r message=FALSE, warning=FALSE, error=FALSE}

dt <- df$Time[2]-df$Time[1] # temporal resolution (s)

paste("Temporal resolution is", dt, "s")

fs <- 1/(dt) # sampling frequency (Hz)

paste("Sampling frequency is", fs, "Hz")

```

With the sampling frequency and window size defined in time, we were able to calculate the number of samples for each window. 

```{r message=FALSE, warning=FALSE, error=FALSE}

n.wnd <- (500/1000)*fs # number of samples for each window

paste(n.wnd, "samples by window")

```

Now, we have to keep in mind that windowing can't exceed the duration of the signal. In addition, the end index of the first window will be the start index of the second window and thus successively until the last window.

```{r message=FALSE, warning=FALSE, error=FALSE}

indx1 <- seq(from = 1,
             to = floor(length(df$Time) - n.wnd),
             by = n.wnd)

print(paste("Start indices:",
            paste(indx1,collapse = ", "),
            collapse = "\n"))

indx2 <- seq(from = 1 + n.wnd,
             to = length(df$Time),
             by = n.wnd)

print(paste("End indices:",
            paste(indx2,collapse = ", "),
            collapse = "\n"))

```

With dygraphs, we can graphically illustrate the windowing performed. To indicate the beginning and end of windows, we can use the ``dyEvent()`` function.

```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center', fig.width=9, fig.height=3.5}

df.filtered <- df %>%
  dplyr::select(Time, EMG2)


dygraph(df.filtered, main = "EMG - channel 2", ylab = "", xlab = "time (s)") %>%
  dyRangeSelector(dateWindow = c(1.95,3.55)) %>%
  dyOptions(colors = "#fb8500") %>%
  dyShading(from = df$Time[indx1[5]], to = df$Time[indx2[5]], color = "#fee44047") %>% 
  dyEvent(c(df$Time[indx1[5]],df$Time[indx2[5]]), c("wnd 5 - start", "wnd 5 - end"), labelLoc = "bottom") %>%
  dyShading(from = df$Time[indx1[6]], to = df$Time[indx2[6]], color = "#fdf4db") %>% 
  dyEvent(c(df$Time[indx1[6]],df$Time[indx2[6]]), c("wnd 6 - start", "wnd 6 - end"), labelLoc = "top") %>%
  dyShading(from = df$Time[indx1[7]], to = df$Time[indx2[7]], color = "#fee44047") %>% 
  dyEvent(c(df$Time[indx1[7]],df$Time[indx2[7]]), c("wnd 7 - start", "wnd 7 - end"), labelLoc = "bottom")

```

## Removing linear and non-linear trends from biomedical signals

```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center', fig.width=9, fig.height=3.5}

dygraph(df.filtered, main = "EMG - channel 2") %>%
  dyRangeSelector() %>%
  dyOptions(colors = "#fb8500")

trend <- vector(mode = "numeric", length = length(indx1))
tm <- vector(mode = "numeric", length = length(indx1))

for(i in 1:length(trend)){
  
  trend[i] <- median(df$EMG2[indx1[i]:indx2[i]])
  tm[i] <- mean(df$Time[indx1[i]:indx2[i]])
  
}

dygraph(data.frame(wnd = 1:length(trend),
                   trend), 
        main = "trend",
        xlab = "window",
        ylab = "amplitude"
) %>% 
  dyRangeSelector() %>%
  dyOptions(colors = "#bf0000") 


trd <- spline(x = tm,
              y = trend,
              xout = df$Time, 
              method = "fmm")$y

df.filtered$trend <- trd


dygraph(df.filtered, main = "EMG and trend", ylab = "", xlab = "time (s)") %>%
  dyRangeSelector() %>%
  dyOptions(colors = c("#fb8500","#bf0000")) 


EMG2r <- data.frame(Time = df.filtered$Time,
                    EMGr = df.filtered$EMG2 - df.filtered$trend)

dygraph(EMG2r, main = "EMG without trends", ylab = "", xlab = "time (s)") %>%
  dyRangeSelector() %>%
  dyOptions(colors = "#e0301e")

```

## Filtering

```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center', fig.width=9, fig.height=3.5}

library(signal)
library(dygraphs)

df.filtered <- df %>% 
  dplyr::select(Time, A2Y)

dygraph(df.filtered,
        xlab = "time (s)", 
        ylab = "linear acceleration (g)") %>%
  dyRangeSelector() %>%
  dyOptions(colors = "#800080")

dt <- df$Time[2]-df$Time[1] # temporal resolution (s)

paste("Temporal resolution is", dt, "s")

fs <- 1/(dt) # sampling frequency (Hz)

paste("Sampling frequency is", fs, "Hz")

fc_l <- 1 # cutoff frequency (Hz)

fc_h <- 7 # cutoff frequency (Hz)

paste("Cutoff frequency [",fc_l,"-",fc_h, "] Hz", sep = "")

n <- 2 # order 

paste("Filter order is", n)

w <- c(fc_l,fc_h)/(fs/2) # digital frequency

paste("Digital frequency is [",
      paste(w, collapse = "-"), "]", sep = "")

bt <- signal::butter(n = n, W = w, type = "pass") # Butterworth filter coefficients

freqz(filt = bt, Fs = fs)

df.filtered$A2Yf <- filtfilt(bt,df.filtered$A2Y)

dygraph(df.filtered, 
        xlab = "time (s)",
        ylab = "linear acceleration (g)") %>%
  dyRangeSelector() %>%
  dyOptions(colors = c("#800080", "black"))

library(psd)

psd.A2Y <- pspectrum(df.filtered$A2Y,
                     verbose = FALSE, 
                     niter=10, AR=TRUE,
                     x.frqsamp=fs,
                     plot=FALSE) # adaptive multitaper spectrum estimation

psd.A2Y <- psd.A2Y[c("freq","spec")] %>% 
  as.data.frame()

psd.A2Yf <- pspectrum(df.filtered$A2Yf, verbose = FALSE, 
                      niter=10, AR=TRUE,
                      x.frqsamp=fs,
                      plot=FALSE) # adaptive multitaper spectrum estimation

psd.A2Yf <- psd.A2Yf[c("freq","spec")] %>% 
  as.data.frame()

```

```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center', fig.width=9, fig.height=1.5}

g <- ggplot()
g <- g + geom_line(data = psd.A2Y,
                   mapping = aes(x = freq,
                                 y = spec),
                   color = "black", 
                   size = 0.75)
g <- g + geom_line(data = psd.A2Yf,
                   mapping = aes(x = freq,
                                 y = spec),
                   color = "red",
                   size = 0.75)
g <- g + theme_minimal()
g <- g + labs(x = "frequency (Hz)",
              y = "energy (g²/Hz)")
g

```

**Now, it is time for action!**

> Choose the signal that has the most evident linear trend. Apply a Butterworth filter to remove this trend.

# Visualisation of features extracted from biomedical signals

$$activity = \sigma^2 = \frac{1}{N-1} \sum_{i=1}^{N}(x_i - \mu)^2$$

$$mobility = \frac{\sigma_d}{\sigma}$$

$$complexity = \frac{\sigma_{dd}/\sigma_d}{\sigma_d/\sigma}$$
```{r message=FALSE, error=FALSE,warning=FALSE, fig.align='center', fig.height=2.5, fig.width=9.0}

dt <- df$Time[2]-df$Time[1] # temporal resolution (s)

paste("Temporal resolution is", dt, "s")

fs <- 1/(dt) # sampling frequency (Hz)

paste("Sampling frequency is", fs, "Hz")

n.wnd <- (1)*fs # number of samples for each window

paste(n.wnd, "samples by window")

indx1 <- seq(from = 1,
             to = floor(length(df$Time) - n.wnd),
             by = n.wnd)

print(paste("Start indices:",
            paste(indx1,collapse = ", "),
            collapse = "\n"))

indx2 <- seq(from = 1 + n.wnd,
             to = length(df$Time),
             by = n.wnd)

print(paste("End indices:",
            paste(indx2,collapse = ", "),
            collapse = "\n"))

v.var <- vector(mode = "numeric", length = length(indx1))
v.mob <- vector(mode = "numeric", length = length(indx1))
v.comp <- vector(mode = "numeric", length = length(indx1))

v.var.f <- vector(mode = "numeric", length = length(indx1))
v.mob.f <- vector(mode = "numeric", length = length(indx1))
v.comp.f <- vector(mode = "numeric", length = length(indx1))


for(i in 1:length(indx1)){
  
  v.var[i] <- var(df.filtered$A2Y[indx1[i]:indx2[i]])
  
  v.mob[i] <- sqrt((var(diff(df.filtered$A2Y[indx1[i]:indx2[i]])))/var(df.filtered$A2Y[indx1[i]:indx2[i]]))
  
  v.comp[i] <- sqrt(var(diff(diff(df.filtered$A2Y[indx1[i]:indx2[i]])))/ var(diff(df.filtered$A2Y[indx1[i]:indx2[i]])))/v.mob[i]

  
  v.var.f[i] <- var(df.filtered$A2Yf[indx1[i]:indx2[i]])
  
  v.mob.f[i] <- sqrt((var(diff(df.filtered$A2Yf[indx1[i]:indx2[i]])))/var(df.filtered$A2Yf[indx1[i]:indx2[i]]))
  
  v.comp.f[i] <- sqrt(var(diff(diff(df.filtered$A2Yf[indx1[i]:indx2[i]])))/ var(diff(df.filtered$A2Yf[indx1[i]:indx2[i]])))/v.mob[i]
  
    
}


features <- data.frame(
  activity = v.var,
  mobility = v.mob,
  complexity = v.comp,
  activity.f = v.var.f,
  mobility.f = v.mob.f,
  complexity.f = v.comp.f
)

library(tidyverse)

df.longer <- pivot_longer(data = features,
                          cols = 1:6,
                          names_to = "feature",
                          values_to = "value")

df.longer$signal <- ifelse(is.na(str_extract(df.longer$feature, ".f")),
  "raw", "filtered")

df.longer$signal <- factor(df.longer$signal, levels = c("raw","filtered"))

df.longer$feature <- str_replace(df.longer$feature,
                                 fixed(".f"),
                                 "")
ggplot(data = df.longer,
       mapping = aes(x = signal, y = value, color = signal)) + 
  geom_boxplot(outlier.size = 0.2,
               outlier.shape = 1) + 
  facet_wrap(~feature, ncol = 3, scales = "free_y") + 
  theme_minimal() + 
  labs(x = "", y = "")


```

**Now, it is time for action!**

> Estimate the root mean square (RMS) of the filtered and non-filtered signals. The RMS should be estimated for windows of 1 second each, without overlapping. Contrast the estimated features through box plots.
